---
title: "colms_model"
author: "Colm Kennedy"
date: "2022-12-09"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Import data
```{r}
library(ggplot2)
library(moderndive)
library(dplyr)
library(skimr)
all <- read.csv('data/all.csv')
head(all)
```


# Transformations

```{r}
allT <- all %>%
  mutate('income'= log(income),
         'gdpcapita'= log(gdpcapita),
         'healthspend'=log(healthspend),
         'co2'=log(co2))
# Merge oceania with Asia
allT <- allT %>%
  mutate('continent' = ifelse(continent == 'Oceania','Asia',continent))
```

# Parallel Model

Dropping: F = 77.3 adj R^2 =0.8804
*healthspend t = -0.034 p = 0.972850 F = 85.85 adj R^2 = 0.8816
*gdpcapita t =0.097 p = 0.923148 F = 96.3 adj R^2 = 0.8827
*water t=-0.919 p=0.360008  F = 108.4 adj R^2 = 0.8829
*continent
*baby2
```{r}
#colms_model <- lm(data = allT, lifeexp ~ childmort + income + gdpcapita + chdperwoman + healthspend + co2 + water + continent + baby2)
colms_modelP <- lm(data = allT, lifeexp ~ childmort + income + co2)
summary(colms_modelP)

```

# Interaction: Childmort by Continent
```{r}
child_model <- lm(data = allT, lifeexp ~ childmort * continent)
summary(child_model)
```

# Interaction: Water by continent
*Europe not significant, combine it with Africa, or drop water from model
```{r}
water_model <- lm(data = allT, lifeexp ~ water * continent)
summary(water_model)
```

# Interaction: income by continent
*Not significant, keep as parallel model
```{r}
income_model <- lm(data = allT, lifeexp ~ income * continent)
summary(income_model)
```

# Interaction: healthspend by continent
*Only America significant, drop var or combine levels
```{r}
health_model <- lm(data = allT, lifeexp ~ healthspend * continent)
summary(health_model)
```

# Interaction: gdpcapita by continent
*Only America significant, drop var or combine levels
```{r}
gdp_model <- lm(data = allT, lifeexp ~ gdpcapita * continent)
summary(gdp_model)
```

# My Model suggestion

```{r}
colms_model <- lm(data = allT, lifeexp ~ income + co2 + childmort * continent)
summary(colms_model)

```
