---
title: "Ethan_model"
author: "Ethan Matta"
date: "2022-12-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
# Set up
```{r}
library(ggplot2)
library(moderndive)
library(dplyr)
library(skimr)
all <- read.csv('data/all.csv')
head(all)

allT <- all %>%
  mutate('income'= log(income),
         'gdpcapita'= log(gdpcapita),
         'healthspend'=log(healthspend),
         'co2'=log(co2))
# Merge oceania with Asia
allT <- allT %>%
  mutate('continent' = ifelse(continent == 'Oceania','Asia',continent))
```

# what will I be analyzing: income, gdpcapita, healthspend, co2, water, child mort, chd per woman, continent, pop, murder
# what will be in the model: income, water, childmort, maybe:  healthspend, gdpcapita, drop: co2, chdperwoman, pop, murder,
# what i decided to keep in analysis: income can be kept but should not have an interaction model with most things
# 

```{r}
cor(allT$income, allT$healthspend, allT$gdpcapita, allT$co2, allT$water, allT$childmort,allT$chdperwoman,  allT$population, allT$murder)

cor(allT$income, allT$healthspend, allT$gdpcapita)
cor(allT[,c("income","gdpcapita", "healthspend", "co2", "water", "childmort", "chdperwoman", "population", "murder")])
```
# testing various models based on correlation values
```{r}
tempmodel <- lm(data = allT, lifeexp ~ income * continent)
get_regression_table(tempmodel)
points<-get_regression_points(tempmodel)
ggplot(data=points, aes(x=income, y=residual)) + geom_point() + geom_hline(yintercept=0)
get_regression_summaries(tempmodel)


tempmodel2 <- lm(data = allT, lifeexp ~ income)
get_regression_table(tempmodel2)
points<-get_regression_points(tempmodel2)
ggplot(data=points, aes(x=income, y=residual)) + geom_point() + geom_hline(yintercept=0)
get_regression_summaries(tempmodel2)

tempmodel3 <- lm(data = allT, lifeexp ~ income*gdpcapita)
get_regression_table(tempmodel3)
points<-get_regression_points(tempmodel3)
ggplot(data=points, aes(x=income, y=residual)) + geom_point() + geom_hline(yintercept=0)
get_regression_summaries(tempmodel3)

tempmodel4 <- lm(data = allT, lifeexp ~ income*murder)
get_regression_table(tempmodel4)
points<-get_regression_points(tempmodel4)
ggplot(data=points, aes(x=income, y=residual)) + geom_point() + geom_hline(yintercept=0)
get_regression_summaries(tempmodel4)

tempmodel4 <- lm(data = allT, lifeexp ~ income*water)
get_regression_table(tempmodel4)
points<-get_regression_points(tempmodel4)
ggplot(data=points, aes(x=income, y=residual)) + geom_point() + geom_hline(yintercept=0)
get_regression_summaries(tempmodel4)

tempmodel4 <- lm(data = allT, lifeexp ~ income*childmort)
get_regression_table(tempmodel4)
points<-get_regression_points(tempmodel4)
ggplot(data=points, aes(x=income, y=residual)) + geom_point() + geom_hline(yintercept=0)
get_regression_summaries(tempmodel4)

tempmodel4 <- lm(data = allT, lifeexp ~ income+water*childmort) # potential model 1
get_regression_table(tempmodel4)
get_regression_summaries(tempmodel4)

```
# testing various models to find significant matches
```{r}
tempmodel4 <- lm(data = allT, lifeexp ~ co2*childmort) # potential model 2
get_regression_table(tempmodel4)
get_regression_summaries(tempmodel4)


tempmodel4 <- lm(data = allT, lifeexp ~ water*childmort*gdpcapita*healthspend*co2) # potential model 3
get_regression_table(tempmodel4)
get_regression_summaries(tempmodel4)

tempmodel4 <- lm(data = allT, lifeexp ~ water*childmort*gdpcapita*co2) # potential model 3
get_regression_table(tempmodel4)
get_regression_summaries(tempmodel4)

tempmodel4 <- lm(data = allT, lifeexp ~ water+co2*gdpcapita) # potential model 3
get_regression_table(tempmodel4)
get_regression_summaries(tempmodel4)


tempmodel4 <- lm(data = allT, lifeexp ~ popdensity*healthspend) # potential model 3
get_regression_table(tempmodel4)
get_regression_summaries(tempmodel4)
```

```{r}
colms_model <- lm(data = allT, lifeexp ~ income + co2 + childmort * continent)
get_regression_table(colms_model)
get_regression_summaries(colms_model)
```
# my final model
# I decided on this model as it is the model with consistently low p-values and a high R_squared
```{r}
Ethans_model <- lm(data=allT, lifeexp~income+co2+water*childmort)
get_regression_table(Ethans_model)
get_regression_summaries(Ethans_model)


```

